---
title: "R loops are slow"
author: "Dmytro Kryvokhyzha"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Create a test matrix
```{r test-matrix}
set.seed(123)
m <- matrix(runif(1000000, max = 100), ncol = 100)
```

Let's test a trivial loop that cal culculates row means:
```{r simple-loop}
loopmean <- function(x){
  v <- c()
  for (i in c(1:dim(x)[1])){
    v[i] <- mean(x[i,])
  }
  return(v)
}
system.time(loopmeanD <- loopmean(m))
```

Adding additonal characters inside the loop slows it down:
```{r loop-brackets}
loopmeanBrackets <- function(x){
  v <- c()
  for (i in c(1:dim(x)[1])){
    v[i] <- mean(((((((((((((((((x[i,])))))))))))))))))
  }
  return(v)
}
system.time(loopmeanBracketsD <- loopmeanBrackets(m))
identical(loopmeanD, loopmeanBracketsD)
```

Processing by columns is little faster:
```{r simple-loop-column}
loopmeanColumn <- function(x){
  v <- c()
  for (i in (1:dim(x)[2])){
    v[i] <- mean(x[,i])
  }
  return(v)
}
tm <- t(m)
system.time(loopmeanColumnD <- loopmeanColumn(tm))
identical(loopmeanD, loopmeanColumnD)
```

Processing is faster when you allocate the memory for the output object.
```{r loop-vertor}
# Vector
vertorloopmean <- function(x){
  v <- vector(length = dim(x)[1])
  for (i in c(1:dim(x)[1])){
    v[i] <- mean(x[i,])
  }
  return(v)
}
system.time(vertorloopmeanD <- vertorloopmean(m))
identical(loopmeanD, vertorloopmeanD)
```

Usually `apply` is faster than `for` loops:
```{r apply-row}
applymean <- function(x){
  v <- apply(m, 1, mean)
  return(v)
}
system.time(applymeanD <- apply(m, 1, mean))
identical(loopmeanD, applymeanD)
```

`apply` by columns is also faster:
```{r apply-column}
applymeanColumn <- function(x){
  v <- apply(tm, 2, mean)
  return(v)
}
system.time(applymeanColumnD <- applymeanColumn(m))
identical(loopmeanD, applymeanColumnD)
```

You can compile *R* function to byte code to make it little faster:
```{r compiled-loop}
library(compiler)

loopmeanCompiled <- cmpfun(loopmean)
loopmeanCompiledT <- system.time(loopmeanCompiledD <- loopmeanCompiled(m))
loopmeanCompiledT
identical(loopmeanD, loopmeanCompiledD)
```

But it is always better to use the built-in functions:
```{r rowMean}
system.time(rowMeanD <- rowMeans(m))
identical(loopmeanD, rowMeanD)
```

If you know `C++`, you can use in inside *R* with [*Rcpp* library](https://cran.r-project.org/web/packages/Rcpp/index.html):

```{Rcpp rcpp-for-loop}
#include <Rcpp.h>
using namespace Rcpp;

//[[Rcpp::export]]
NumericVector cRowMeans(NumericMatrix x) {
 int nrows = x.nrow();   
 NumericVector v(nrows);
 for (int i = 0; i < nrows; i++){
   v[i] = mean(x.row(i));
 }
 return v;
}
```

It improves the performance damatically:
```{r Rcpp_loop}
system.time(cRowMeansD <- cRowMeans(m))
identical(loopmeanD, cRowMeansD)
```


But it is the fastest to use the built-in function that call *C* code directly.
```{r loopParallel, message=FALSE}
library(doParallel)
registerDoParallel(cores=12)
loopParallelT <- system.time(loopParallelD <- foreach(i=1:dim(m)[1], .combine=c) %dopar% mean(m[i,]))
loopParallelT
identical(loopmeanD, loopParallelD)
```


## Benchmark everything
```{r benchmark-loops}
library(microbenchmark)
mbm <- microbenchmark("LoopRowExtraBrackets" = loopmeanBrackets(m),
                      "ApplyRow" = applymean(m),
                      "ApplyColumn" = applymeanColumn(tm),
                      "LoopRow" = loopmean(m),
                      "LoopRowCompiled" = loopmeanCompiled(m),
                      "LoopColumn" = loopmeanColumn(tm),
                      "LoopRowToVector" = vertorloopmean(m),
                      "RowLoopC"= cRowMeans(m),
                      "Built-in_rowMeans" = rowMeans(m),
                      check = 'equal', times=100)

mbm
```

Visualize
```{r benchmark-plot, message=FALSE, fig.width=8, fig.height=5}
library(ggplot2)
autoplot(mbm)
```

Alternative benchmark:
```{r benchmark-loops-alternative}
library(rbenchmark)
bm <- benchmark("LoopRowExtraBrackets" = loopmeanBrackets(m),
                "ApplyRow" = applymean(m),
                "ApplyColumn" = applymeanColumn(tm),
                "LoopRowCompiled" = loopmeanCompiled(m),
                "LoopRow" = loopmean(m),
                "LoopColumn" = loopmeanColumn(tm),
                "LoopRowToVector" = vertorloopmean(m),
                "Built-in_rowMeans" = rowMeans(m),
                columns=c("test", "elapsed", "relative", "user.self", "sys.self"),
                order="relative", replications=100)
bm
#sbarplot(bm$user.self, names.arg = bm$test)
```
